Estimating Uncertainty for Tipping Points with Deep Learning

### 2022-05-09 manuscript/manuscript.Rmd

Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@kao, @lyu, @Du].
Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance.
Memory and attention are proposed as advantageous for time series forecasting since these methods enable models to learn long term dependencies [@kao, @attentionisall].
Encoder-decoder architectures have been shown to excel on sequence to sequence tasks [@learningphase].
To investigate these assertions on the challenging class of tipping point problems, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.
We will examine LSTM and GRU models, which are both memory-based networks, Transformers, which use attention with an encoder-decoder, and Block LSTM's, which employ LSTM's in an encoder-decoder framework.

@@

A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@cite].
We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
One proposed reason for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@modelstability]. 
Another likely reason for this instability is overfitting to the training set.
Overfitting is a fundamental issue in deep learning, for which there is no universal solution [@ying].
A common method to reduce overfitting is to perform K-fold cross validation [@ying], but this approach cannot be effectively employed when there is one or few time series in the training set.
To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 models that were trained with different initialization seeds.
We found this simple ensemble method to be an effective way to improve the generalizability for each ML model.

>>>>

Over the past decade, deep learning has become very popular for a broad range of challenging time series prediction problems [@kao, @lyu, @du].
Across these studies, it has become established that using memory or attention-based neural networks, and an encoder-decoder architecture is crucial for improving forecasting performance -- see @kao, @lyu and @du for more details on these concepts.
Memory and attention are proposed as advantageous for time series forecasting since these methods enable models to learn long term dependencies [@kao, @attentionisall].
Encoder-decoder architectures have been shown to excel on sequence to sequence tasks [@learningphase].
To investigate these assertions on the challenging class of tipping point problems, we examine four models that use a combination of memory, attention and encoder-decoder network architectures.
We will examine LSTM and GRU models, which are both memory-based networks, Transformers, which use attention with an encoder-decoder, and Block LSTM's, which employ LSTM's in an encoder-decoder framework.

@@

A major nuisance with deep learning methods is their instability to hyperparameters and initialization seeds [@cite].
We found that for the same set of hyperparameters, we could produce starkly different forecasts if we trained two models with different initialization seeds [@figure_].
One proposed reason for this instability is that machine learning models often get stuck on the local optima of loss surfaces [@modelstability]. 
Another likely reason is that machine learning models commonly overfit the training data.
Overfitting is a fundamental issue in deep learning, for which there is no universal solution [@ying].
One frequently cited method to reduce overfitting is K-fold cross validation [@ying], but this approach cannot be effectively employed when there is one or few time series in the training set.
To remedy the instability problem, we use an ensemble model, wherein each ML forecast is the union of forecasts from 5 individual models that were trained with different initialization seeds.
We found this simple ensemble method to be an effective way to improve generalizability.
